---
title: "GAPMINDER: Life Expectancy Comparisons"
author: "Magdalena Krysztofiak"
date: "2026-03-25"
output: html_document
---

FREE DATA FROM WORLD BANK VIA GAPMINDER.ORG, CC-BY LICENSE
<https://www.gapminder.org/>

```{r setup, include = FALSE}
knitr::opts_chunk$set( #global chunk options
	echo = FALSE,
	message = FALSE,
	warning = FALSE
) #show only output, supress warning and messages
library(readr)
library(tidyverse)
#install.packages("effsize")
library(effectsize)
```

```{r data, include = FALSE}
gapminder <- read_csv("gapminder.csv")
```

### Filter 1952

Let's select data only for the year 1952.

Options:

- "==" equal to, 
- "!=" not equal to 
- "<" less than, 
- "<=" less or equal to, 
- ">" more, 
- ">=" more or equal to 
- "&" and, 
- "|" or
- %in% membership operator, is x in y

```{r filter-gapminder}
gapminder %>%
  filter(year == 1952) -> gapminder_52
```

### Explore

Let's explore distrubution of lifeExp (numeric variable) in our data for 1952

####  Histogram

First, histogram...

```{r histogram}
ggplot(gapminder_52, aes(x = lifeExp)) + geom_histogram()
```

...we can adjust the number (bins) or the width of the bins (binwidth)
Binwidth is measured in the units of x variable.

```{r histogram-bins}
ggplot(gapminder_52, aes(x = lifeExp)) + 
  geom_histogram(binwidth = 1)
```

Practice 1:
How to show the distrubution of lifeExp on each continent?

```{r practice-1}
ggplot(gapminder_52, aes(x = lifeExp)) + 
  geom_histogram() +
  facet_wrap(~continent)
```

####  Barplot

Why is it not working?

```{r barplot-1}
ggplot(gapminder, aes(x = lifeExp)) + geom_bar()
```

Observations are binned 
and frequency in each bin is mostly = 1

```{r group-size}
gapminder %>% group_by(lifeExp) %>% summarise(n())
```

What if we round the values?

```{r barplot-2}
gapminder %>%
  mutate(lifeExp_rounded = round(lifeExp)) %>%
  ggplot(aes(x = lifeExp_rounded)) + geom_bar()
```

###  Density plot

We can see the shape of the distribution,
it's not affected by the number of bins.

```{r density-plot}
ggplot(gapminder_52, aes(x = lifeExp)) + 
  geom_density(adjust = 0.5)
```

### Compare life expectancy between Asia and Europe in 2002.

Practice 2: 
Make a new df with rows for Asia or Europe and the year 2002.

Options:

- "==" equal to, 
- "!=" not equal to 
- "<" less than, 
- "<=" less or equal to, 
- ">" more, 
- ">=" more or equal to 
- "&" and, 
- "|" or
- %in% membership operator, is x in y

```{r filter-continents-year}
gapminder %>% 
  filter(year == 2002, 
           continent %in% c("Asia","Europe")) -> gapminder_filtered
```

We will compare life expectancy between Asia and Europe in 2002.

What would be the best way to visualize the difference 
between Asia and Europe in life expectancy?

```{r visualize}
gapminder_filtered %>% ggplot(aes(continent, lifeExp)) + 
  geom_boxplot()
```

Summary of life expectancy for Asia and Europe:
*How can we get summary statistics (n, mean, sd, range) for each continent?*

```{r summarize}
gapminder_filtered %>%
  group_by(continent) %>%
  summarise(n = n(), 
            median = median(lifeExp),
            mean = mean(lifeExp), 
            sd = sd(lifeExp), 
            min = min(lifeExp), 
            max = max(lifeExp))
```
We will compare life expectancy with Wilcoxon test
because the variable is not normally distrubuted.

We can check the effect size (useful for broader interpretation in this case).

```{r compare}
wilcox.test(lifeExp ~ continent, data = gapminder_filtered)
rank_biserial(lifeExp ~ continent, data = gapminder_filtered)
```
