---
title: "GAPMINDER: Life Expectancy Comparisons"
author: "Magdalena Krysztofiak"
date: "2026-03-25"
output: html_document
---

FREE DATA FROM WORLD BANK VIA GAPMINDER.ORG, CC-BY LICENSE
<https://www.gapminder.org/>

```{r setup, include = FALSE}
knitr::opts_chunk$set( #global chunk options
	echo = FALSE,
	message = FALSE,
	warning = FALSE
) #show only output, supress warning and messages
library(readr)
library(tidyverse)
#install.packages("effsize")
library(effectsize)
```

```{r data, include = FALSE}
read_csv("gapminder.csv") -> gapminder
```

### Filter 1952

Let's select data only for the year 1952.

Options:

- "==" equal to, 
- "!=" not equal to 
- "<" less than, 
- "<=" less or equal to, 
- ">" more, 
- ">=" more or equal to 
- "&" and, 
- "|" or
- %in% membership operator, is x in y

```{r filter-gapminder}

```

### Explore

Let's explore distrubution of lifeExp (numeric variable)
in our data for 1952

####  Histogram

First, histogram...

```{r histogram}

```

...we can adjust the number (bins) or the width of the bins (binwidth)
Binwidth is measured in the units of x variable.

```{r histogram-bins}

```

Practice 1:
How to show the distrubution of lifeExp on each continent?

```{r practice-1}

```

####  Barplot

Why is it not working?

```{r barplot-1}

```

Observations are binned 
and frequency in each bin is mostly = 1

```{r group-size}

```

What if we round the values?

```{r barplot-2}

```

###  Density plot

We can see the shape of the distribution,
it's not affected by the number of bins.

```{r density-plot}

```

### Compare life expectancy between Asia and Europe in 2002.

Practice 2: 
Make a new df with rows for Asia or Europe and the year 2002.

Options:

- "==" equal to, 
- "!=" not equal to 
- "<" less than, 
- "<=" less or equal to, 
- ">" more, 
- ">=" more or equal to 
- "&" and, 
- "|" or
- %in% membership operator, is x in y

```{r filter-continents-year}

```

We will compare life expectancy between Asia and Europe in 2002.

What would be the best way to visualize the difference 
between Asia and Europe in life expectancy?

```{r visualize}

```

Summary of life expectancy for Asia and Europe:
*How can we get summary statistics (n, mean, sd, range) for each continent?*

```{r summarize}

```

We will compare life expectancy with Wilcoxon test
because the variable is not normally distrubuted.

We can check the effect size (useful for broader interpretation in this case).

```{r compare}

```
